---
title: "Prediction Assignment Writeup"
author: "Chris Little"
date: "06/17/2015"
output:
  html_document: null
  pdf_document: null
---
```{r results="hide", echo=FALSE, message=FALSE, cache=TRUE}
library(knitr)

# set options
opts_chunk$set(cache = TRUE)

# load libraries
library(caret)
library(randomForest)

# set a seed value
# set.seed(as.integer(Sys.time()))
set.seed(1337)

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

setwd("~/coursework/predmachlearn")
```

# Data Clean-up

First, we load the training & testing data from disk.
```{r cache=TRUE}
pml_train <- read.csv("./data/pml-training.csv", na.strings = c("", "NA", "#DIV/0!", '""'))
pml_test <- read.csv("./data/pml-testing.csv", na.strings = c("", "NA", "#DIV/0!", '""'))
```

Then, we clean the data frames up by removing irrelevent columns. Columns 1-7 contain non-sensor data.
```{r}
pml_train <- pml_train[, 8:ncol(pml_train)]
pml_test <- pml_test[, 8:ncol(pml_test)]
```

Columns with near zero variance won't be useful as discriminating predictors, so they are removed.
```{r}
nzv_cols <- nzv(pml_train)
pml_train <- pml_train[, -nzv_cols]
pml_test <- pml_test[, -nzv_cols]
rm(nzv_cols)
```

Columns that are entirely NAs in the test set obviously can't be used for prediction, so they are removed.
```{r}
no_na_cols <- !as.logical(colSums(is.na(pml_test)))
pml_train <- pml_train[,  no_na_cols]
pml_test <- pml_test[, no_na_cols]
rm(no_na_cols)
```

```{r cache=TRUE, results="hide", echo=FALSE, message=FALSE}
# train a random forests predictor
# rf_fit <- randomForest(classe ~ ., data=pml_train)
```

Next, using random forests with PCA-preprocessing, a predictor is trained from the training set. All of the sensor data is used to build the model that predicts classe. 5-fold cross-validation is performed by the trainer.
```{r cache=TRUE, message=FALSE}
tc <- trainControl(method = "cv", number = 5)
rf_fit <- train(classe ~ ., data = pml_train,  preProcess = c("pca"),  method = "rf",
                  prox = TRUE, trControl = tc)
```


```{r cache=TRUE, results="hide", echo=FALSE, message=FALSE}
#rf_pred <- predict(rf_fit, pml_valid[, -ncol(pml_valid)])
#confusionMatrix(rf_pred, pml_valid$classe)
```

With a well-trained model, we perform prediction on the test set:
```{r}
answers <- predict(rf_fit, pml_test[, -ncol(pml_test)])
pml_write_files(answers)
```

# References

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. *Qualitative Activity Recognition of Weight Lifting Exercises.* Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. http://groupware.les.inf.puc-rio.br/har