---
title: "Prediction Assignment Writeup"
author: "Chris Little"
date: "06/17/2015"
output:
  pdf_document: null
  html_document: null
---

# Introduction

This is a writeup illustrating the application of machine learning (specifically 5-fold cross-validated random forests) to data from [Velloso, et al. 2013] in order to predict activity type from sensor data.

```{r results="hide", echo=FALSE, message=FALSE, cache=TRUE}
library(knitr)

# set options
opts_chunk$set(cache = TRUE)

# load libraries
library(caret)
library(randomForest)

# set a seed value
# set.seed(as.integer(Sys.time()))
set.seed(1337)

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

setwd("~/coursework/predmachlearn")
```

# Data Clean-up

First, we load the training & testing data from disk.
```{r cache=TRUE}
pml_train <- read.csv("./data/pml-training.csv", na.strings = c("", "NA", "#DIV/0!", '""'))
pml_test <- read.csv("./data/pml-testing.csv", na.strings = c("", "NA", "#DIV/0!", '""'))
```

Then, we clean the data frames up by removing irrelevent columns. Columns 1-7 contain non-sensor data.
```{r cache=TRUE}
pml_train <- pml_train[, 8:ncol(pml_train)]
pml_test <- pml_test[, 8:ncol(pml_test)]
```

Columns with near zero variance won't be useful as discriminating predictors, so they are removed.
```{r cache=TRUE}
nzv_cols <- nzv(pml_train)
pml_train <- pml_train[, -nzv_cols]
pml_test <- pml_test[, -nzv_cols]
```

Columns that are entirely NAs in the test set obviously can't be used for prediction, so they are removed.
```{r cache=TRUE}
no_na_cols <- !as.logical(colSums(is.na(pml_test)))
pml_train <- pml_train[,  no_na_cols]
pml_test <- pml_test[, no_na_cols]
```

```{r cache=TRUE, results="hide", echo=FALSE, message=FALSE}
# train a random forests predictor
# rf_fit <- randomForest(classe ~ ., data=pml_train)
```

# Model Training and Cross-Validation

Next, using random forests with PCA-preprocessing, a predictor is trained from the training set. All of the sensor data is used to build the model that predicts classe. 5-fold cross-validation is performed by the trainer. (This model takes a long time to train, so it is serialized to disk and only trained from the data if it is absent from disk.)
```{r message=FALSE}
if (!file.exists("rf_fit.rds")) {
  tc <- trainControl(method = "cv", number = 5)
  rf_fit <- train(classe ~ ., data = pml_train,  preProcess = c("pca"),  method = "rf",
                  prox = TRUE, trControl = tc)
  saveRDS(rf_fit, file = "rf_fit.rds")
  rf_fit
} else {
  readRDS(rf_fit, file = "rf_fit.rds")
}
```

With the selected model, accuracy is 0.981 (so we expect an error rate of 0.019) and $\kappa$ is 0.976.

A cross-validated confusion matrix of the model's predictions illustrates its good fit:
```{r cache=TRUE}
confusionMatrix(rf_fit)
```

```{r cache=TRUE, results="hide", echo=FALSE, message=FALSE}
#rf_pred <- predict(rf_fit, pml_valid[, -ncol(pml_valid)])
#confusionMatrix(rf_pred, pml_valid$classe)
```

```{r cache=TRUE, results="hide", echo=FALSE, message=FALSE}
# With a well-trained model, we perform prediction on the test set:
answers <- predict(rf_fit, pml_test[, -ncol(pml_test)])
pml_write_files(answers)
```

# References

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. *Qualitative Activity Recognition of Weight Lifting Exercises.* Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. http://groupware.les.inf.puc-rio.br/har

